ANCOVA Comparison Simulations: Discrete Outcomes

### 2017-10-13 results/results.Rmd

We reanalyzed a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD). 
Patients were treated at eight sites in two different countries. 
At each site, patients were randomly assigned one of two treatments. 
Patients were observed for a week before receiving treatment and for a week after receiving treatment. 
On each of the fourteen days of observation, patients responded to a survey about their heartburn, regurgitation, and dyspepsia frequency and severity. 
These endpoints were measured on a discrete scale. 
There were several additional endpoints, such as daily regurgitation, daily “hrdq”, and daily dyspepsia, calculated from the survey measures. 
Daily “hrdq” was the primary endpoint. 
To reduce day-to-day variation, we averaged the measures from each week to obtain two observations per patient, one pre-treatment and one post-treatment.

@@

@@

```
Figure \ref{fig:hist_continuous_vars} shows the distribution of the final outcome measures between the two treatment groups.
Based on these distributions, we would expect to detect a difference in outcomes between the treatments for the `daily_heart`, `daily_hrdq`, and `heart_freq` measures.

@@

Table 1 shows the correlation between baseline and outcome for each variable.
Correlations range between 0.548 and 0.696.
This suggests that using the difference in outcome and baseline as the dependent variable in the following models might yield higher power than controlling for baseline as a covariate.

@@

summarytabc <- xtable(baseline_outcome_corr, 
             digits = 3,
             align = paste0(c("r|", rep("p{1in}", 1)), 
                            collapse = ""),
             caption = "Correlation between baseline and outcome, for each continuous endpoint.",
             label = "tab:clinical_corr")
print(summarytabc, include.rownames = TRUE)
```

@@

@@

```{r hrdq_parametric, results = "asis"}
# method 0: parametric ANCOVA
lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
print(xtable(summary(aov(lm1))), include.rownames = TRUE)
daily_hrdq$lm1_resid <- residuals(lm1)

lm2 <- lm(difference ~ tr + factor(SITEID), data = daily_hrdq)
print(xtable(summary(aov(lm2))), include.rownames = TRUE)
daily_hrdq$lm2_resid <- residuals(lm2)
```

@@

@@

### Linear model residuals

@@

By construction of the linear model, the correlation between the baseline measure and the residuals is zero.
However, we would like to check formally that $\varepsilon$ is independent baseline measure for each site, using the residuals as a proxy for $\varepsilon$.
To do so, we conduct independent permutation tests within each stratum using the Pearson correlation as the statistic.
Table 3 shows the results of this test: while the correlation between baseline and residuals appears high, it is not statistically significant at the 5\% level in any stratum.
The Fisher's combined $p$-value for the 8 strata is `r round(fisher_combined_corr, 2)`, so there is no evidence that the independence assumption is violated.

```{r residual_plots, echo = FALSE, message = FALSE, fig.align = "center", fig.height = 5, fig.cap = "Residual plots of the linear regression of daily hrdq on treatment and covariates. These are done to check that treatment and covariates are uncorrelated with the regression errors for the Freedman and Lane style permutation test."}
p11 <- daily_hrdq %>% 
  select(-Outcome, -difference) %>% 
  melt(id.vars = c("SUBJID", "Baseline", "SITEID")) %>%
  filter(variable == "lm1_resid") %>% 
  ggplot(aes(x = Baseline, y = as.numeric(value))) +
  geom_point() + 
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Baseline") + ylab("Residual") + 
  ggtitle("Raw Outcome Model") +
  theme_bw()
p21 <- daily_hrdq %>% 
  select(-Baseline, -Outcome, -difference) %>%  
  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
  filter(variable == "lm1_resid") %>% 
  ggplot(aes(x = factor(SITEID), y = as.numeric(value))) +
  geom_boxplot() +  
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Site ID") + ylab("Residual") + 
  ggtitle("Raw Outcome Model") +
  theme_bw()
p22 <- daily_hrdq %>% 
  select(-Baseline, -Outcome, -difference) %>%  
  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
  filter(variable == "lm2_resid") %>% 
  ggplot(aes(x = factor(SITEID), y = as.numeric(value))) +
  geom_boxplot() +  
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Site ID") + ylab("Residual") +
  ggtitle("Differenced Model") +
  theme_bw()
grid.arrange( p21, p11, p22, layout_matrix = rbind(c(1, 2), c(3, NA)))
```

@@

We restrict our attention to the model using treatment measurement and controlling for the baseline, as this is what RB's original analysis does.
We run this procedure for all seven continuous endpoints.
Table 5 shows the results.

```{r results_table_make}
set.seed(919547773) # Generated on Random.org Timestamp: 2016-11-09 15:39:29 UTC

@@

>>>>

We reanalyzed a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD). 
Patients were treated at eight sites in two different countries. 
At each site, patients were randomly assigned one of two treatments. 
Patients were observed for a week before receiving treatment and for a week after receiving treatment. 
On each of the fourteen days of observation, patients responded to a survey about their heartburn, regurgitation, and dyspepsia frequency and severity. 
These endpoints were measured on a discrete scale. 
There were several additional endpoints, such as daily regurgitation, daily “hrdq”, and daily dyspepsia, calculated from the survey measures. 
Daily “hrdq” was the primary endpoint. 
To reduce day-to-day variation, we averaged the measures from each week to obtain two observations per patient, one pre-treatment and one post-treatment.
These averages are no longer discrete ordinal values, but they are not continuous either: they can only take a finite set of values.

@@

We show that the data may violate the assumptions for the parametric ANCOVA.
In particular, the outcome variables and the linear regression residuals are skewed to the right, and there is evidence of heteroskedasticity.
The small $p$-values that come from the parametric analysis may therefore be unreliable.
However, the permutation tests yield only slightly larger $p$-values.
The correspondence between parametric and permutation test results gives some confidence that the results are stable and reliable.

\newpage

@@

```
Figure \ref{fig:hist_continuous_vars} shows the distribution of the final outcome measures between the two treatment groups.
Table 1 shows the means of the distributions.
Based on these summary statistics, we would expect to detect a difference in outcomes between the treatments for the `daily_heart`, `daily_hrdq`, and `heart_freq` measures.

@@

```{r summary_tab, results = "asis"}
outcome_summary <- clinical %>% 
  group_by(tr) %>% 
  summarise(mean(daily_heart),
    mean(daily_regurg), 
    mean(daily_dysp),
    mean(daily_hrdq), 
    mean(heart_freq),
    mean(regurg_freq), 
    mean(dysp_freq)) %>%
  select(-tr) %>%
  round(2) %>% t()
outcome_sds <- clinical %>% 
  group_by(tr) %>% 
  summarise(sd(daily_heart),
    sd(daily_regurg), 
    sd(daily_dysp),
    sd(daily_hrdq), 
    sd(heart_freq),
    sd(regurg_freq), 
    sd(dysp_freq)) %>%
  select(-tr) %>%
  round(2) %>% t()
outcome_summary <- cbind(outcome_summary[, 1],
                         outcome_sds[, 1],
                         outcome_summary[, 2],
                         outcome_sds[, 2])
colnames(outcome_summary) <- c("Mean A", "SD A", "Mean B", "SD B")
rownames(outcome_summary) <- continuous_vars
print(xtable(outcome_summary, caption = "Mean and standard deviation of each continuous outcome in groups A and B."), include.rownames = TRUE)

Table 2 shows the correlation between baseline and outcome for each variable.
Correlations range between 0.548 and 0.696.
This suggests that using the difference in outcome and baseline as the dependent variable in the following models might yield higher power than controlling for baseline as a covariate.

@@

summarytabc <- xtable(baseline_outcome_corr, 
             digits = 3,
             align = paste0(c("r|", rep("p{1in}", 1)), 
                            collapse = ""),
             caption = "Correlation between baseline and outcome, for each continuous endpoint.",
             label = "tab:clinical_corr")
print(summarytabc, include.rownames = TRUE)
```
\newpage

@@

\newpage

@@

```{r hrdq_parametric, results = "asis"}
# method 0: parametric ANCOVA
lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
daily_hrdq$lm1_resid <- residuals(lm1)
daily_hrdq$lm1_fitted <- fitted(lm1)

lm2 <- lm(difference ~ tr + factor(SITEID), data = daily_hrdq)
daily_hrdq$lm2_resid <- residuals(lm2)
daily_hrdq$lm2_fitted <- fitted(lm2)
```

@@

```{r hrdq_parametric_tab, results = "asis"}
print(xtable(summary(aov(lm1))), include.rownames = TRUE)
print(xtable(summary(aov(lm2))), include.rownames = TRUE)
```

```{r residual_plot_lm_hrdq, fig.align="center", fig.height = 3, fig.cap = "Residual plots for the ANCOVA model of daily hrdq."}
p01 <- daily_hrdq %>% 
  ggplot(aes(x = lm1_fitted, y = lm1_resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted Values") + ylab("Residual") + 
  ggtitle("Raw Outcome Model") +
  theme_bw()
p02 <- daily_hrdq %>% 
  ggplot(aes(x = lm2_fitted, y = lm2_resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted Values") + ylab("Residual") + 
  ggtitle("Differenced Model") +
  theme_bw()
grid.arrange(p02, p01, nrow=1)
```

Figure \ref{fig:residual_plot_lm_hrdq} shows the fitted values versus residuals for the two ANCOVA models.
The residuals of the raw outcome model appear to increase in magnitude with the fitted value, casting doubt on the assumption that the error variance is constant.
The plot reveals that the differenced model only yields a few predicted values, but does not seem to indicate heteroskedasticity.

@@

### Freedman-Lane test of residuals

@@

By construction of the linear model, the correlation between the baseline measure and the residuals is zero.
However, we would like to check formally that $\varepsilon$ is independent baseline measure for each site, using the residuals as a proxy for $\varepsilon$.
To do so, we conduct independent permutation tests within each stratum using the Pearson correlation as the statistic.
Table 5 shows the results of this test: while the correlation between baseline and residuals appears high, it is not statistically significant at the 5\% level in any stratum.
The Fisher's combined $p$-value for the 8 strata is `r round(fisher_combined_corr, 2)`, so there is no evidence that the independence assumption is violated.

```{r residual_plots, echo = FALSE, message = FALSE, fig.align = "center", fig.height = 5, fig.cap = "Residual plots of the linear regression of daily hrdq on treatment and covariates. These are done to check that treatment and covariates are uncorrelated with the regression errors for the Freedman and Lane style permutation test."}

p11 <- daily_hrdq %>% 
  select(-Outcome, -difference) %>% 
  melt(id.vars = c("SUBJID", "Baseline", "SITEID")) %>%
  filter(variable == "lm1_resid") %>% 
  ggplot(aes(x = Baseline, y = as.numeric(value))) +
  geom_point() + 
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Baseline") + ylab("Residual") + 
  ggtitle("Raw Outcome Model") +
  theme_bw()
p21 <- daily_hrdq %>% 
  select(-Baseline, -Outcome, -difference) %>%  
  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
  filter(variable == "lm1_resid") %>% 
  ggplot(aes(x = factor(SITEID), y = as.numeric(value))) +
  geom_boxplot() +  
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Site ID") + ylab("Residual") + 
  ggtitle("Raw Outcome Model") +
  theme_bw()
p22 <- daily_hrdq %>% 
  select(-Baseline, -Outcome, -difference) %>%  
  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
  filter(variable == "lm2_resid") %>% 
  ggplot(aes(x = factor(SITEID), y = as.numeric(value))) +
  geom_boxplot() +  
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Site ID") + ylab("Residual") +
  ggtitle("Differenced Model") +
  theme_bw()
grid.arrange( p21, p11, p22, layout_matrix = rbind(c(1, 2), c(3, NA)))
```

@@

We restrict our attention to the model using treatment measurement and controlling for the baseline, as this is what RB's original analysis does.
We run this procedure for all seven continuous endpoints.
Table 6 shows the results.

Figure \ref{fig:residual_plots_lm} displays the residuals versus fitted values for the linear model using each response variable.
The figures cast doubt on the assumption of constant error variance in the linear model: they indicate that the variance of the residuals increases with the fitted values and the distribution of residuals may have some right skew.
The assumption of normally distributed errors with constant variance is important for the parametric analysis of the ANCOVA model, but does not apply to the permutation tests.
Across all tests, the parametric ANCOVA $p$-value is smaller than the linear model permutation test $p$-values.
Perhaps this is partly due to the heteroskedasticity: the permutation tests make no distributional assumptions, so may be less sensitive to large outcomes or residuals.
```{r results_table_make}
set.seed(919547773) # Generated on Random.org Timestamp: 2016-11-09 15:39:29 UTC

@@

```{r residual_plots_lm, fig.align = "center", fig.cap = "Residual plots for the linear regression of outcome on baseline, treatment, and stratum ID for each response variable."}

pplot <- vector("list", 7)
for(i in seq_along(continuous_vars)){
  col <- continuous_vars[i]
  tmpdata <- reshape_data(col)
  lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = tmpdata)
  tmpdata <- tmpdata %>% mutate("Residuals" = residuals(lm1), "Fitted" = fitted(lm1))
  pplot[[i]] <- ggplot(tmpdata, aes(x = Fitted, y = Residuals)) + geom_point() + ggtitle(continuous_vars[i]) +
  theme_bw()
}
do.call(grid.arrange, c(pplot, nrow= 2))
```

### 2017-10-13 simulations/discrete_outcomes.Rmd

@@

```{r vary_effect_size_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
```

@@

```{r vary_effect_size_het_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_het_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
```
Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
The pattern here is nearly identical to Figure 2 in the continuous outcome simulations, so we leave out further comments.

```{r plot_for_ms, include=FALSE}
p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
  ylim(0, 0.95 ) +
  theme(legend.position="bottom") + 
  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
  report_theme +
  ggtitle("Constant Treatment \n Effect")

p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
    ylim(0, 0.95) +
    report_theme +
    ggtitle("Heterogeneous Treatment \n Effect")

>>>>

plot_power_ratio_gamma <- function(powermat_list, gamma_vec, alpha, title){
  gamma_power_alpha <- t(sapply(powermat_list, function(x) x[floor(alpha*nrow(x)),]))
  colnames(gamma_power_alpha) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
  gamma_power_alpha <- apply(gamma_power_alpha, 2, function(x) x/gamma_power_alpha[,"ANCOVA"])
  melt(gamma_power_alpha, value.name="Power") %>%
  mutate("Method" = Var2) %>%
  mutate("Effect" = rep(gamma_vec, 5)) %>% 
  filter(Method != "Differenced Permutation") %>%
  filter(Method != "ANCOVA") %>%
  ggplot(aes(x = Effect, y = Power)) +
  geom_line(aes(color = Method))+
  xlab("Effect size") +
  ylab("Relative Power") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    title = element_text(size = 16),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    strip.text.x = element_text(size = 12)
  ) 
}
```

```{r vary_effect_size_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
plot_power_ratio_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect")

@@

@@

```{r vary_effect_size_het_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_het_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect")

plot_power_ratio_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Constant Treatment Effect")

```
Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
The pattern here is nearly identical to Figure 2 in the continuous outcome simulations, so we leave out further comments.

```{r plot_for_ms, include=FALSE}
p1 <- plot_power_ratio_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
  ylim(0, 1.2 ) +
  theme(legend.position="bottom") + 
  guides(color=guide_legend(nrow=1,byrow=TRUE)) +
  report_theme +
  ggtitle("Constant Treatment \n Effect")

p2 <- plot_power_ratio_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
  ylim(0, 1.2 ) +
    report_theme +
    ggtitle("Heterogeneous Treatment \n Effect")
